---
title: "Workshop 8 TEMPLATE"
subtitle: "Friday 9 AM"
author: "Mila Matich"
date: "2026-05-22"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data", "Valliere_etal_EcoApps_Data.xlsx"),
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plants_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col,
  "Stipa pulchra" = stipul_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw()) # will be used for all future plots
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# storing data as a new objects (creating new object from original frame)
drought_exp_clean <- drought_exp |> 
  # cleaning column names (making them lowercase with underscore)
  clean_names() |> 
  # creating new columns to store binomial species name 
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # creating new columns with full water treatment name
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  ))  |> 
  # reordering species level, lotus scoparius as our reference 
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # reordering water treatment levels 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting columns of interest to display 
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  
```{r}
#| label: checking-structure 

str(drought_exp_clean)

```

Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: 10-rows

slice_sample(
  drought_exp_clean, # data frame to display 
  n = 10) # selecting 10 rows of data to display 
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

```{r}
#| label: making-scatter-plot

ggplot(data = drought_exp_clean, 
       # x: specific leaf area 
       # y = total plant mass 
       mapping = aes(x = sla,
                     y = total_g)) + 
  # geom point for scatter plot 
  geom_point(color = stipul_col) + 
  # relabeling x-axis and y-axis 
  labs(x = "Specific Leaf Area (cm\U00B2/g)", # how to do a superscript 
       y = "Total Plant Mass (g)",
       title = "As Specific Lead Area Increases, Total Mass also Increases") + 
  theme_set()
```

## What are the differences in total mass between water treatments?

```{r}
#| label: jitterplot-visualization 
 
# calculating the summary for mean point 
drought_exp_summary <- drought_exp_clean |> 
  group_by(water_treatment) |> # grouping by treatment
  summarize(mean = mean(total_g)) |> 
  ungroup()

# generating the plot     
ggplot(data = drought_exp_clean,
       # x-axis = treatment 
       # y-axis = total mass
       # color by treatment 
       mapping = aes(x = water_treatment,
                     y = total_g,
                     color = water_treatment)) + 
  # jitter represent idnidivual plants 
  geom_jitter(width = 0.1, # narrows the jitter
              height = 0) + # y-axis position represents the value
  # adding mean point
  geom_point(data = drought_exp_summary,
             mapping = aes(x = water_treatment,
                           y = mean),
             color = "red") + 
  # adding custom colors 
  scale_color_manual(values = c("Well watered" = ww_col,
                                "Drought Stressed" = ds_col)) + 
  # lebeling axis 
  labs(x = "Water Treatment",
       y = "Total Mass (g)",
       title = "On average, well watered plants have higher mass\nthan drought stressed plants") + # \n to create a line break 
  # setting + adjusting themes 
  theme_set() + 
  theme(legend.position = "none",
        plot.title.position = "plot") # title on top of y-axis  
```

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.


## What are the differences in total mass between species?

```{r}
#| label: new-jitter-mass-vs-species

# making new summary 
drought_exp_summary1 <- drought_exp_clean |> 
  group_by(species_name) |> # grouping by treatment
  summarize(mean = mean(total_g)) |> 
  ungroup()
  
# base layer = ggplot 
ggplot(data = drought_exp_clean,
       # x-axis = species 
       # y-axis = total mass
       # color by species 
       mapping = aes(x = species_name,
                     y = total_g,
                     color = species_name)) + 
  # first layer: geom jitter
  geom_jitter(width = 0.1,
              height = 0,
              alpha = 0.4) + 
  # using costum colors 
  scale_color_manual( values = plants_colors) +
  # wrapping x-axis text 
  scale_x_discrete(labels = label_wrap(10)) +
  # adding point for the mean per species 
  geom_point(data = drought_exp_summary1,
             mapping = aes(x = species_name,
                           y = mean),
             color = "red") + 
  # labeling x and y + title 
  labs(x = "Species Name",
       y = "Total Mass (g)",
       title = "Mean plant mass differs across species") + 
  # setting + adjusting the theme to take out legend
  theme_set() + 
  theme(legend.position = "none") 

```


# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model

model0 <- lm(
  total_g = 1, # formula: no predictors 
  data = drought_exp_clean # data = clean data fram 
)

# model 1: all predictors

model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula: all predictors 
  data = drought_exp_clean # data = clean data frame
)

# model 2: SLA and water treatment

model2 <- lm(
  total_g ~ sla + water_treatment, # formula: only SLA and treatment
  data = drought_exp_clean # data = clean data frame
)

# model 3: SLA and species

model3 <- lm(
  total_g ~ sla + species_name, # formula: only SLA and species
  data = drought_exp_clean # data = clean data frame
)

# model 4: water treatment and species

model4 <- lm(
  total_g ~ water_treatment + species_name, # formula: only treatment and species
  data = drought_exp_clean # data = clean data frame
)

# model 5: SLA

model5 <- lm(
  total_g ~ sla, # formula: only sla
  data = drought_exp_clean # data = clean data frame
)

# model 6: water treatment

model6 <- lm(
  total_g ~ water_treatment, # formula: only water treatment
  data = drought_exp_clean # data = clean data frame
)

# model 7: species

model7 <- lm(
  total_g ~ species_name, # formula: only species
  data = drought_exp_clean # data = clean data frame
)

```

## Model diagnostics

```{r}
#| label: check-diagnostics 

# displaying plots in 2x2 grid
par(mfrow = c(2,2))
# generating diagnostic plots for model 2
plot(model4)
```

- residuals appear normally distributed 
- residuals appear homoscedastic (they have constant variance)
- no clear outliers in data set influencing model coefficient estimates 

## Model selection

```{r}
#| label: calculating-AIC

# calculating AIC
AICc(
  model0,
  model1, 
  model2, 
  model3,
  model4, 
  model5, 
  model6, 
  model7
) |> 
  # rearranging output to be in order of lowest to highest AIC value
  arrange(AICc) 

```

Insert code to display the summary for the best model. 
```{r}
#| label: model-summary 
# displaying the model summary 
summary(model4)
```

- '(intercept)': represents the model prediction for plant mass for lotus scoparius in the drought stressed treatment (because lotus and drought stressed are the reference for both groups/variables (know this because they do not show up in the output))
- 'water_treatmentWell watered': model prediction for the difference in Lotus scoparius mass between the well watered and drought stressed treatment 
- 'species_namePestemon centranthifolius': model prediction for the difference between the Pestemon mass in the well watered treatment and Lotus mass in the well watered treatment 
- 'species_nameEschscholzia californica ': model prediction for the difference between the Eschscholzia mass in the well watered treatment and lotus mass in the well watered treatment

## Stop and think: what does this model mean?  

**What is the best model?**

Our best model is model 4, water treatment and species (as indicates by the smallest AICc value) as predictors.

**How much variation in the response (total mass, in grams) does this model explain?**

72.6 percent of the variation in plant mass is explained by a model with both of these predictors.

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

- Treatment influences plant mass 
- Species influences plant mass

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions 

# new prediction object 
preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  # renaming columns
  rename(water_treatment = x,
         species_name = group)
```


# 5. Final figure

```{r}
#| label: final-figure 

ggplot(data = drought_exp_clean, 
       mapping = aes(x = water_treatment, 
                     y = total_g,
                     color = water_treatment)) + 
  geom_jitter(height = 0, 
              width = 0.1,
              shape = 21,
              alpha = 0.8) + 
  facet_wrap(~ species_name) + 
  geom_pointrange(data = preds,
                  mapping = aes(x = water_treatment, 
                               y = predicted, 
                               ymin = conf.low, 
                               ymax = conf.high)) + 
  theme(legend.position = "none") + 
  labs(x = "Water Treatment",
       y = "Total Plant Mass (g)")
```



# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







